Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-03-10 draft.tex
Replies to Fred's comments

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to arbitrary decay or scattering
processes \fred{Not really arbitrary, it only works with orthogonal
  basis functions}. We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrizations or approximations.

@@

We continue with basic definitions that pertain to angular observables, and our
results in the subsequent sections. Let $\thvec$ denote the set of all
angles, and let $\nuvec$ denote the set of all
other non-angular kinematic variables needed to fully specify the
final state of the process under study. For example, $\nuvec$ may
include invariant masses or center-of-mass energies. We define an
angular observable $S_i$ as a coefficient in the probability density
function (PDF), $P(\nuvec, \thvec)$, of the process by
\begin{align}
    \label{eq:def-P}
    P(\nuvec, \thvec) \equiv \sum_i S_i(\nuvec) \times f_i(\thvec)\,.
\end{align}
Here, the dependence on the decay angles $\thvec$ has been
explicitly factored out in terms of the angular functions
$\{f_i(\thvec)\}$. We assume there exists a dual basis of functions
$\{\dual{f}_i(\thvec)\}$ such that the orthonormality relations
\begin{equation}
    \label{eq:def-ortho-rel}
    \int_\Omega \rmdx{\vec{\theta}} \dual{f}_i(\thvec) f_j(\thvec)  = \delta_{ij}
\end{equation}
hold with $\Omega$ representing the full angular phase space relevant
to the process.  For particle decays, $P$ is generally expressed in
terms of the fully differential decay width,
\begin{align}
    \label{eq:def-P-decay}
    P(\nuvec, \thvec) \equiv \frac{1}{\Gamma}\frac{\rmdx{^{2}\Gamma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where $\Gamma$ is the total decay width. For a scattering process, one can similarly use
\begin{align}
    \label{eq:def-P-scattering}
    P(\nuvec, \thvec) \equiv \frac{1}{\sigma}\frac{\rmdx{^{2}\sigma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where the total cross section $\sigma$ is used for the
normalization. Since the determination of the total decay width or
total cross section can be quite difficult, we emphasize that
different normalizations for $P$ can be used.  For instance, the total
decay width (or cross section) of the process of interest can be
replaced by the corresponding quantity of a control-channel
process. This change of normalization is equivalent to a linear rescaling of
the angular observables $\lbrace S_i\rbrace$; thus ratios or similar suitable combinations
of the angular observables are not affected by this change of normalization.\\

Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. It turns out to be much more convenient to work
in the basis of Legendre polynomials but the statistical properties
are equivalent for both approaches \fred{I'm confused. What are the two approaches? Do we consider Legendre polynomials as an orthonormal basis or not?} provided that one replace $f_i \to
\dual{f}_i$ appropriately. Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:btokstarll}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular functions. In general, $\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through \ref{app:btokstarll}.\\

@@

Integration over the non-angular variables yields
\begin{equation}
    \langle S_i\rangle
    \equiv \int \rmdx{\vec\nu} S_i(\nuvec)
    = \int \rmdx{\vec \nu} \left[\int_{\Omega} \rmdx{\vec \theta} P(\nuvec,\thvec) \dual{f}_i(\thvec) \right].
\end{equation}
\fred{Why is there $\Omega$ but nothing for $\nuvec$?}
The remainder of this section describes the method of moments, in
which we replace the analytical integration by Monte Carlo (MC)
estimates.  The central tenet of MC integration is the fact that the
expectation value $E_P[g]$ of some function $g(x)$ under the
probability density $P(x)$,
\begin{equation}
    E_P[g] \equiv \int \rmdx{x} P(x) g(x),
\end{equation}
can be approximated by the consistent and unbiased
estimator $\est{E_P[g]}$~\cite[sec. 8.2]{James:2006zz}
\begin{equation}
    \label{eq:mc-id}
    E_P[g] \to \widehat{E_P[g]} \equiv \frac{1}{N} \sum_{n=1}^N g(x^{(n)}) \,,\,    x^{(n)} \sim P
\end{equation}
due to the strong law of large numbers for $N \to \infty$, assuming
that the variates $x^{(n)}$, $n = 1, \dots, N$, are distributed as
$P$.
Throughout this letter we denote all MC estimators with a wide hat.\\

@@

\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
  smaller than those from the method of moments  \fred{Is this true, Marcin?}.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

@@

For the decay at hand, we consider the partial-wave expansion for the angular observables (see \cite{Lee:1992ih} and \refapp{partial-waves})
\begin{equation}
    S_k(\vec{\nu}) \equiv \frac{I_k(\nuvec)}{4\pi \Gamma)}\,,\quad\text{with}\quad \Gamma = I_1(\nuvec) - \frac{I_2(\nuvec)}{3}\,,
\end{equation}
\fred{It not clever to use $k$ as an index and momentum variable}
which reads
\begin{equation}
    \label{eq:expansion}
    S_{k}(\vec{\nu},\cos\theta_2) \equiv \sum_{j} \frac{1}{n_{l,|m|}} S_{k,l}(\vec{\nu}) p_{l}^{(|m|)}(\cos\theta_2)\sin\theta_2\,.
\end{equation}
The normalization factor $n_{l,|m|}$ is defined in \refeq{legendre-scalar-product}. Within our
example we have (cf. also \refapp{btokstarll} and \cite{Lee:1992ih})
\begin{equation}
    |m| = \begin{cases}
        0\,, & k = 1,2,6\,,\\
        1\,, & k = 4,5,7,8\,,\\
        2\,, & k = 3,9
    \end{cases}\,.
\end{equation}

@@

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates.
For an experimental analysis, one would now turn to the determination of the distribution of $\vecest{S}$
as a function of the detector setup. This would involve the determination of both $M$ and $\vecest{Q}$ for a number of
detector configurations, and subsequent profiling or marginalization.
While such considerations of any detector simulation are beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples \fred{Before or after thinning?} (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find that the error on the mean
of each matrix element is smaller than $6\cdot 10^{-4}$. \add{This suggests, that the so-induced
systematic error can be driven below any statistical uncertainty.}

>>>>

Our initial motivation for studying what we wish to call the
\emph{method of moments} is the determination of angular observables
in the rare FCNC-mediated decay $\bar{B}\to \bar{K}^*(\to
\bar{K}\pi)\ell^+\ell^-$. However, the method we describe in the
following is general and applies to \add{all decay or scattering
processes that can be formulated in terms of an orthogonal basis of angular functions.}
\fred{Not really arbitrary, it only works with orthogonal
basis functions.} \danny{Is this OK?}
We find a previous work \cite{Dighe:1998vk} that
advocates this method chiefly for the determination of angular
observables in non-leptonic $B$ decays but also mentions the
applicability to semileptonic decays. Our aim is to improve upon this
previous work by studying the uncertainties that are introduced by
mismodeling of the angular distribution, and by working out a recipe
to determine and unfold detector effects. The latter is crucial for
the application of the method to real data. We show that
the method of moments has several major advantages over the usual approach
based on likelihood fits:
\begin{enumerate}
    \item Likelihood fits have convergence problems for a small number of
        events, and can require reparametrizations and/or approximations
        for a successful fit to the signal PDF. \add{As an example,
        see the LHCb analysis of \dots}.
        The method of moments does not require reparametrizations or approximations.

@@

We continue with basic definitions that pertain to angular observables, and our
results in the subsequent sections. Let $\thvec$ denote the set of all
angles, and let $\nuvec$ denote the set of all
other non-angular kinematic variables needed to fully specify the
final state of the process under study. For example, $\nuvec$ may
include invariant masses or center-of-mass energies. We define an
angular observable $S_i$ as a coefficient in the probability density
function (PDF), $P(\nuvec, \thvec)$, of the process by
\begin{align}
    \label{eq:def-P}
    P(\nuvec, \thvec) \equiv \sum_i S_i(\nuvec) \times f_i(\thvec)\,.
\end{align}
Here, the dependence on the decay angles $\thvec$ has been
explicitly factored out in terms of the angular functions
$\{f_i(\thvec)\}$. We assume there exists a dual basis of functions
$\{\dual{f}_i(\thvec)\}$ such that the orthonormality relations
\begin{equation}
    \label{eq:def-ortho-rel}
    \int_\Omega \rmdx{\vec{\theta}} \dual{f}_i(\thvec) f_j(\thvec)  = \delta_{ij}
\end{equation}
hold with $\Omega$ representing the full angular phase space relevant
to the process.  For particle decays, $P$ is generally expressed in
terms of the fully differential decay width,
\begin{align}
    \label{eq:def-P-decay}
    P(\nuvec, \thvec) \equiv \frac{1}{\Gamma}\frac{\rmdx{^{2}\Gamma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where $\Gamma$ is the total decay width. For a scattering process, one can similarly use
\begin{align}
    \label{eq:def-P-scattering}
    P(\nuvec, \thvec) \equiv \frac{1}{\sigma}\frac{\rmdx{^{2}\sigma}}{\rmdx{\vec\nu} \rmdx{\vec\theta}}\,,
\end{align}
where the total cross section $\sigma$ is used for the
normalization. Since the determination of the total decay width or
total cross section can be quite difficult, we emphasize that
different normalizations for $P$ can be used.  For instance, the total
decay width (or cross section) of the process of interest can be
replaced by the corresponding quantity of a control-channel
process. This change of normalization is equivalent to a linear rescaling of
the angular observables $\lbrace S_i\rbrace$; thus ratios or similar suitable combinations
of the angular observables are not affected by \add{a} change of normalization.\\

Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. \add{We find it convenient to work
in the basis of Legendre polynomials, which is not self-dual.
However, the statistical properties of the method of moments are equivalent for both self-dual and non-self-dual
bases, provided that one replace $f_i \to \dual{f}_i$ appropriately.}
\fred{I'm confused. What are the two approaches? Do we consider Legendre polynomials as an orthonormal basis or not?}
Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:btokstarll}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular functions
\add{, and $\Omega$ represents the entire angular phase space.}. In general, $\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through \ref{app:btokstarll}.\\

@@

Integration over the non-angular variables yields
\begin{equation}
    \langle S_i\rangle
    \equiv \int \rmdx{\vec\nu} S_i(\nuvec)
    = \int \rmdx{\vec \nu} \left[\int_{\Omega} \rmdx{\vec \theta} P(\nuvec,\thvec) \dual{f}_i(\thvec) \right].
\end{equation}
\fred{Why is there $\Omega$ but nothing for $\nuvec$?}
\danny{a) intoduced $\Omega$ one equation earlier, b) we always need
and integration of the \emph{entire} angular phase space; the integration
over the non-angular phase space is irrelevant to the discussion.}
The remainder of this section describes the method of moments, in
which we replace the analytical integration by Monte Carlo (MC)
estimates.  The central tenet of MC integration is the fact that the
expectation value $E_P[g]$ of some function $g(x)$ under the
probability density $P(x)$,
\begin{equation}
    E_P[g] \equiv \int \rmdx{x} P(x) g(x),
\end{equation}
can be approximated by the consistent and unbiased
estimator $\est{E_P[g]}$~\cite[sec. 8.2]{James:2006zz}
\begin{equation}
    \label{eq:mc-id}
    E_P[g] \to \widehat{E_P[g]} \equiv \frac{1}{N} \sum_{n=1}^N g(x^{(n)}) \,,\,    x^{(n)} \sim P
\end{equation}
due to the strong law of large numbers for $N \to \infty$, assuming
that the variates $x^{(n)}$, $n = 1, \dots, N$, are distributed as
$P$.
Throughout this letter we denote all MC estimators with a wide hat.\\

@@

\item The estimate is unbiased for any $N$. In contrast, the
  maximum-likelihood estimate has a bias of order
  $1/N$~\cite{Cox:1968}. In practice, one should keep in mind the
  bias-variance trade-off: it is a well known phenomenon that removing
  the bias usually leads to an increase in variance of the sampling
  distribution of the estimator and vice versa~\cite[sec. 7.3]{James:2006zz}. From a
  Bayesian decision-theory point of view, both contribute similarly to
  the expected loss associated with deciding on just one value of the
  unknown parameter. \add{One should therefore not prefer the method of moments
  over likelihood fits just because the former reduces}
  the bias~\cite[sections 13.8,17.2]{jaynes:2003}. In fact, for
  the results discussed below in \refsec{numerics}, the likelihood
  fits --- if they converge --- produce uncertainties \add{(10--15)\%}
      smaller than those from the method of moments  \fred{Is this true, Marcin?} \danny{Marcin already confirmed!}.
\item The approximate multivariate Gaussian distribution of $\vecest{S}$
    allows easier and more \add{precise} transfer of the information in the
  data to interested theorists for more accurate fits of
  standard-model and new-physics parameters \cite{Altmannshofer:2013foa,
Descotes-Genon:2013wba,Beaujean:2013soa},
  or for more precise predictions of optimized observables; see e.g.,
  \cite{Egede:2008uy,Egede:2010zc,Bobeth:2010wg,Becirevic:2011bp,
    Bobeth:2012vn,Matias:2012xw,DescotesGenon:2012zf} for definitions
  of such optimized observables in $B\to K^*\ell^+\ell^-$ decays,
  \cite{Faller:2013dwa} for application to the decay $B\to
  \pi\pi\ell^-\bar\nu_\ell$, and \cite{Boeer:2014xx} for observables
  in $\Lambda_b\to\Lambda(\to N\pi)\ell^+\ell^-$). While the
  likelihood also approaches a multivariate Gaussian as $N \to
  \infty$, the two methods differ in their utility as input for
  theorists if $\widehat{\vec{S}}$ is not well
  inside the physical region. For example, suppose there are two
  angular observables that are constrained to a triangular region by
  phase-space or unitarity arguments as
  \begin{equation}
    \label{eq:constr-ex}
    |S_1| + S_2 \le 1, \, S_1 \in [-1,1],\, S_2 \in [0,1] \,.
  \end{equation}
  It may (and often does) happen in practice that $\widehat{\vec{S}}$
  is close or even outside the allowed region such that a significant
  part of the probability mass covers unphysical values. In a Bayesian
  fit, one would take $\mathcal{N}(\vecest{S} | \vec{S},
  \widehat{\Sigma})$ as the sampling distribution of the ``data'', and
  simply set a uniform prior on the triangle in the $\vec{S}$ plane
  defined by \refeq{constr-ex} to have a well defined problem. This
  could be trivially combined with other independent information in a
  global fit. Someone with a different physics model might have to
  consider a
  different physical region, and could incorporate it just as easily.\\

@@

For the decay at hand, we consider the partial-wave expansion for the angular observables (see \cite{Lee:1992ih} and \refapp{partial-waves})
\begin{equation}
    S_k(\vec{\nu}) \equiv \frac{I_k(\nuvec)}{4\pi \Gamma)}\,,\quad\text{with}\quad \Gamma = I_1(\nuvec) - \frac{I_2(\nuvec)}{3}\,,
\end{equation}
\fred{It not clever to use $k$ as an index and momentum variable}
\danny{I disagree. We use the same notation as in the literature, and people
should be able to distinguish these based on context.}
which reads
\begin{equation}
    \label{eq:expansion}
    S_{k}(\vec{\nu},\cos\theta_2) \equiv \sum_{j} \frac{1}{n_{l,|m|}} S_{k,l}(\vec{\nu}) p_{l}^{(|m|)}(\cos\theta_2)\sin\theta_2\,.
\end{equation}
The normalization factor $n_{l,|m|}$ is defined in \refeq{legendre-scalar-product}. Within our
example we have (cf. also \refapp{btokstarll} and \cite{Lee:1992ih})
\begin{equation}
    |m| = \begin{cases}
        0\,, & k = 1,2,6\,,\\
        1\,, & k = 4,5,7,8\,,\\
        2\,, & k = 3,9
    \end{cases}\,.
\end{equation}

@@

The accuracy of the unfolding process as outlined above critically depends on both the accuracy
of the detector simulation, as well as the uncertainties induced by the MC estimates.
For an experimental analysis, one would now turn to the determination of the distribution of $\vecest{S}$
as a function of the detector setup. This would involve the determination of both $M$ and $\vecest{Q}$ for a number of
detector configurations, and subsequent profiling or marginalization.
While such considerations of any detector simulation are beyond the scope of this work, we can, however,
comment on the MC-induced uncertainties. As usual, one needs to find a balance between
compute time and accuracy. For a uniangular distribution and a total of $6\cdot 10^6$ MC
samples \fred{Before or after thinning?}\danny{No thinning at all!} (i.e., $2 \cdot 10^6$ MC samples per row of $M_{ij}$), we find that the error on the mean
of each matrix element is smaller than $6\cdot 10^{-4}$. \add{This suggests, that the so-induced
systematic error can be driven below any statistical uncertainty.}
